---
title: "Read in DNA methylation data, and pre-filter to normal, prefrontal data"
subtitle: "Read and pre-filter ROSMAP data"
author:
  - David Lukacsovich^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::html_document:
    highlight: breezedark
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Set Up

## Load Libraries

```{r library}
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(doParallel)
  if (Sys.info()["sysname"] == "Windows") {
    library(parallel)
  }
  
  library(GEOquery)
  library(Biobase)
  library(stringr)
  library(plyr)
  library(downloader)
  library(stats)
  library(ggplot2)
  
  source("../R/CorticalClock.r", local=TRUE)
})

base_dir <- "~/TBL Dropbox/David Lukacsovich"
code_dir <- file.path(base_dir, "AD-Aging-brain-sample-analysis/code")
dataset_dir <- file.path(code_dir, "DATASETS", "Rosmap")
download_dir <- file.path(dataset_dir, "01_download")
idat_dir <- file.path(dataset_dir, "02_idat_data")
compile_dir <- file.path(dataset_dir, "03_sample_data")
reference_dir <- file.path(code_dir, "Reference_Files")
```

# Get Phenotype Data

## Read Phenotype Files

```{r read_pheno_files}
# read phenotype files
df_1 <- read.csv(file.path(
  reference_dir, "ROSMAP_clinical_4-8-2024.csv"
)) %>%
  dplyr::select(
    "projid",
    "Study",
    "msex",
    "educ",
    "race",
    "spanish",
    "apoe_genotype",
    "age_at_visit_max",
    "age_first_ad_dx",
    "age_death",
    "cts_mmse30_first_ad_dx",
    "cts_mmse30_lv",
    "pmi",
    "braaksc",
    "ceradsc",
    "cogdx"
  ) %>%
  dplyr::distinct()
df_2 <- read.table(file.path(
  reference_dir, "ROSMAP_arrayMethylation_covariates.tsv"
), header = TRUE) %>%
  dplyr::select(
    "Sample",
    "Sentrix_ID",
    "Sentrix_Position",
    "Sample_Plate",
    "Sample_Well",
    "Sample_Group",
    "batch"
  ) %>%
  dplyr::distinct()
df_3 <- read.csv(file.path(
  reference_dir, "ROSMAP_IDkey.csv"
)) %>%
  dplyr::select(Sample = "mwas_id", "projid") %>%
  dplyr::distinct()
```

## Get List of Existing idat Files

```{r get_idat_list}
df_idat <- list.files(
  idat_dir,
  pattern = "*_Grn.idat",
  all.files = FALSE,
  full.names = FALSE,
  recursive = TRUE
) %>%
  stringr::str_split_fixed(pattern = "_", n=3) %>%
  data.frame() %>%
  dplyr::select(sentrix_id = "X1", slide = "X2") %>%
  dplyr::mutate(sentrix_id = gsub("^.*/", "", .data$sentrix_id)) %>%
  dplyr::mutate(label = paste0(.data$sentrix_id, "_", .data$slide))
```

## Merge Phenotype Files

```{r crete_pheno}
pheno_df <- df_1 %>%
  dplyr::inner_join(df_3, by = "projid") %>%
  dplyr::inner_join(df_2, by = "Sample") %>%
  dplyr::filter(paste0(.data$Sentrix_ID, "_", .data$Sentrix_Position) %in% df_idat$label) %>%
  dplyr::filter(!is.na(.data$braaksc)) %>%
  dplyr::filter(.data$race == 1) %>%
  dplyr::filter(.data$spanish != 1)

write.csv(pheno_df, file.path(download_dir, "full_pheno.csv"))
saveRDS(pheno_df, file.path(download_dir, "full_pheno.RDS"))
```

## Filter Phenotype Columns

```{r filter_pheno}
pheno_df <- pheno_df %>%
  dplyr::mutate(braaksc = as.integer(as.numeric(.data$braaksc))) %>%
  dplyr::mutate(stage3 = ifelse(
    .data$braaksc < 3, "0-2",
    ifelse(.data$braaksc < 5, "3-4", "5-6")
  )) %>%
  dplyr::mutate(Sentrix = paste0(.data$Sentrix_ID, "_", .data$Sentrix_Position)) %>%
  dplyr::mutate(Slide = .data$Sentrix_ID) %>%
  dplyr::mutate(sex = ifelse(.data$msex == 1, "MALE", "FEMALE")) %>%
  dplyr::mutate(
    age_death = ifelse(
      is.na(.data$age_death),
      .data$age_at_visit_max,
      .data$age_death
    )
  ) %>%
  dplyr::mutate(
    age_death = as.numeric(ifelse(.data$age_death == "90+", "90", .data$age_death))
  ) %>%
  dplyr::mutate(msex = ifelse(.data$msex == 1, "MALE", "FEMALE")) %>%
  dplyr::mutate(ceradsc = as.integer(.data$ceradsc)) %>%
  dplyr::select(
    "Sample",
    "Sentrix_ID", 
    "Sentrix_Position",
    "Sample_Plate",
    "Sample_Well",
    "Sample_Group",
    "batch",
    "Sentrix",
    "Slide", 
    "projid", 
    "Study", 
    "msex",
    "educ", 
    "race",
    "spanish",
    "apoe_genotype",
    "age_at_visit_max",
    "age_first_ad_dx",
    "age_death",
    "cts_mmse30_first_ad_dx",
    "cts_mmse30_lv",
    "pmi",
    "braaksc",
    "ceradsc",
    "cogdx",
    "stage3",
    "sex"
  ) %>%
  dplyr::distinct() %>%
  dplyr::mutate(Sentrix_ID = as.character(.data$Sentrix_ID))
```

# Filter Phenotype Data

## Filter by Age and PMI

```{r get_old}
pheno_df <- pheno_df %>%
  dplyr::filter(
    .data$age_death >= 65
  ) %>%
  dplyr::filter(!is.na(.data$pmi)) %>%
  dplyr::filter(.data$pmi <= 72)
```

## Filter to Normal Data

```{r get_normal}
pheno_df <- pheno_df %>%
  dplyr::filter(.data$stage3 %in% "0-2")
```

# Collect and Filter Assay Data

## Read in idat Data

```{r read_idat}
# read idat data
assay_data <- minfi::read.metharray.exp(
  base = idat_dir,
  recursive = TRUE,
  verbose = FALSE,
  force = TRUE
)

# adjust sample names
df_names <- stringr::str_split_fixed(colnames(assay_data), pattern = "_", n = 2)
colnames(df_names) <- c("Sentrix_ID", "Sentrix_Position")
df_names <- df_names %>%
  as.data.frame() %>%
  dplyr::left_join(
  pheno_df, by = c("Sentrix_ID", "Sentrix_Position")
) %>%
  tidyr::replace_na(list(Sample = "Missing"))
colnames(assay_data) <- df_names$Sample
```

## Filter to Matching Phenotype Data

```{r filter_idat}
assay_data <- assay_data[,pheno_df$Sample]
```

# Predict Age

## Get Beta Values

```{r age_beta}
suppressMessages({
  beta_data <- minfi::getBeta(assay_data)
})
```

## Summarize Missing Cortical Probe Values

```{r coef_summary}
brain_coef <- read.csv(file.path(reference_dir, "brain_coef.csv"), row.names = 1)
beta_coef <- beta_data[row.names(beta_data) %in% brain_coef$probe, ]
beta_missing <- is.na(beta_coef)

df_missing <- data.frame(
  probes = c(
    "CorticalClock probe count (N)",
    "CorticalClock probes in data (N)",
    "Sample count (N)",
    "Probes with Missing Values (N)",
    "Samples with Missing Values (N)",
    "Total Missing Values (N)"
  ),
  value = c(
    nrow(brain_coef),
    nrow(beta_coef),
    ncol(beta_coef),
    sum(rowSums(beta_missing) > 0),
    sum(colSums(beta_missing) > 0),
    sum(beta_missing)
  )
)

df_missing %>%
  DT::datatable(
    filter = 'top',
    style = "bootstrap",
    extensions = 'Buttons',
    options = list(
      scrollX = TRUE, 
      dom = 'Bfrtip',
      buttons = I('colvis'),
      keys = TRUE, 
      pageLength = 10
    ), 
    rownames = FALSE,
    caption = "Number of NAs in CorticalClock Probe Beta Values"
  )
```

## Run Prediction

```{r}
pheno <- CorticalClock(
  beta_data,
  pheno_df,
  paste0(reference_dir, "/"),
  "Sample",
  "age_death"
)

pheno_df$age_prediction <- as.numeric(pheno$brainpred)
```

# Save

```{r save, eval = TRUE}
saveRDS(assay_data, file.path(compile_dir, "organized_assay.RDS"))
saveRDS(pheno_df, file.path(compile_dir, "organized_pheno.RDS"))
write.csv(pheno_df, file.path(compile_dir, "organized_pheno.csv"))
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()
```
</details>