---
title: "Download DNA methylation data, and pre-filter to normal, prefrontal data"
subtitle: "Download and pre-filter BDR data"
author:
  - David Lukacsovich^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::html_document:
    highlight: breezedark
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Set Up

## Load Libraries

```{r library}
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(doParallel)
  if (Sys.info()["sysname"] == "Windows") {
    library(parallel)
  }
  
  library(GEOquery)
  library(Biobase)
  library(stringr)
  library(plyr)
  library(downloader)
  library(stats)
  library(ggplot2)
  
  source("../R/CorticalClock.r", local=TRUE)
})

base_dir <- "~/TBL Dropbox/David Lukacsovich"
analysis_dir <- file.path(base_dir, "AD-Aging-brain-sample-analysis")
code_dir <- file.path(analysis_dir, "code")
dataset_dir <- file.path(code_dir, "DATASETS", "BDR")
download_dir <- file.path(dataset_dir, "01_download")
idat_dir <- file.path(dataset_dir, "02_idat_data")
compile_dir <- file.path(dataset_dir, "03_sample_data")
reference_dir <- file.path(code_dir, "Reference_Files")
result_dir <- file.path(analysis_dir, "analysis-results")
```

# Download Data

## Get Phenotype Data

```{r download_pheno}
geo_id <- "GSE197305"

# Get Phenotype
geo_data <- GEOquery::getGEO(
  GEO = geo_id,
  destdir = download_dir
)

pheno_df <- Biobase::pData(
  geo_data[[paste0(geo_id, "_series_matrix.txt.gz")]]
) %>%
  data.frame()

# Get sample target .idat file information
supp_columns <- c(
  colnames(pheno_df)[stringr::str_starts(colnames(pheno_df),
                                         "supplementary_file")],
  "geo_accession"
)
supp_df <- pheno_df[,supp_columns]

# save data
saveRDS(pheno_df, file.path(download_dir, "full_pheno.RDS"))
saveRDS(supp_df, file.path(download_dir, "full_supplementary.RDS"))
write.csv(pheno_df, file.path(download_dir, "full_pheno.csv"))
```

## Get idat Data

```{r download_idat, eval = FALSE}
files_ls <- dplyr::select(supp_df, -"geo_accession") %>%
  unlist() %>%
  unname() %>%
  unique()

plyr::a_ply(files_ls, .margins = 1, .fun = function(url) {
  output_name <- file.path(idat_dir, basename(url))
  output_unzip <- stringr::str_sub(output_name, 1, nchar(output_name) - 3)
  if(!file.exists(output_name) & !file.exists(output_unzip)) {
    downloader::download(url, output_name)
    GEOquery::gunzip(output_name)
  }
}, .progress = "time")
```

# Filter Phenotype Data

## Filter to Cortical Data

```{r filter_cortical}
pheno_df <- read.csv(file.path(download_dir, "full_pheno.csv"), row.names = 1) %>%
  dplyr::filter(.data$brain.region.ch1 == "Prefrontal")
```

## Cluster Data

### Run Clustering

```{r cluster_cluster}
df_k <- pheno_df %>%
  dplyr::select(
    neun = "cell.proportion.neun..ch1",
    sox10 = "cell.proportion.sox10..ch1",
    negative = "cell.proportion.double.negative.ch1"
  ) %>% dplyr::mutate(
    neun = as.numeric(.data$neun),
    sox10 = as.numeric(.data$sox10),
    negative = as.numeric(.data$negative)
  )

# run clustering
set.seed(1022023)
k_value <- 3
cluster_results <- stats::kmeans(df_k, centers = k_value)
df_k$clustering <- cluster_results$cluster
```

### Label Clusters

```{r cluster_label}
# get sox10 proportion means by cluster
df_stats <- df_k %>%
  dplyr::group_by(.data$clustering) %>%
  dplyr::summarise(ave_sox = mean(.data$sox10)) %>%
  dplyr::arrange(.data$ave_sox) %>%
  dplyr::mutate(braak = c("0-II", "III-IV", "V-VI"))

# merge to original data
df_k <- df_k %>%
  dplyr::left_join(df_stats, by = "clustering")
pheno_df$braak <- df_k$braak
df_k$`Braak stage` <- df_k$braak
```

### Plot Clustering Results

```{r cluster_plot}
p1 <- ggplot2::ggplot(
  df_k,
  mapping = ggplot2::aes(x = neun, y = sox10, color = `Braak stage`)
) +
  ggplot2::geom_point(size = 0.5) +
  ggplot2::ylab("SOX10+") +
  ggplot2::xlab("NeuN+") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  ) +
  ggplot2::theme(legend.position = c(0.85, 0.82))

p2 <- ggplot2::ggplot(
  df_k,
  mapping = ggplot2::aes(x = negative, y = sox10, color = `Braak stage`)
) +
  ggplot2::geom_point(size = 0.5) +
  ggplot2::ylab("SOX10+") +
  ggplot2::xlab("NeuN-/SOX10-") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  ) +
  ggplot2::theme(legend.position = c(0.85, 0.82))

p3 <- ggplot2::ggplot(
  df_k,
  mapping = ggplot2::aes(x = neun, y = negative, color = `Braak stage`)
) +
  ggplot2::geom_point(size = 0.5) +
  ggplot2::xlab("NeuN+") +
  ggplot2::ylab("NeuN-/SOX10-") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  ) +
  ggplot2::theme(legend.position = c(0.85, 0.82))

pdf(
  file.path(result_dir, "meta_analysis", "BDR_Braaks_Plots.pdf"),
  width = 5,
  height = 5
)
p1
p2
p3
dev.off()

pdf(
  file.path(result_dir, "meta_analysis", "BDR_Braaks_Plots_Side.pdf"),
  width = 12,
  height = 4
)
gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
dev.off()

gridExtra::grid.arrange(p1, p2, p3, ncol = 2)
```

## Filter to Normal Data

```{r filter_braak}
pheno_df <- pheno_df %>%
  dplyr::filter(.data$braak == "0-II")
```

## Rename and Filter Phenotype Columns

```{r filter_columns}
pheno_df <- pheno_df %>%
  tidyr::separate(
    "source_name_ch1",
    into = c("sentrix_id", "slide"),
    sep = "_",
    remove = TRUE
  ) %>%
  dplyr::mutate(age = NA) %>%
  dplyr::select(
    sample = "geo_accession",
    "sentrix_id",
    "slide",
    "age",
    sex = "gender.ch1",
    stage = "braak",
    region = "brain.region.ch1"
  ) %>%
  dplyr::mutate(
    sex = ifelse(.data$sex == "female", "FEMALE", .data$sex),
    sex = ifelse(.data$sex == "male", "MALE", .data$sex),
    age = as.numeric(.data$age)
  )
```

# Collect and Filter Assay Data

## Read in idat Data

```{r read_idat}
# read idat data
assay_data <- minfi::read.metharray.exp(
  base = idat_dir,
  recursive = TRUE,
  verbose = FALSE,
  force = TRUE
)

# adjust sample names
colnames(assay_data) <- as.character(
  stringr::str_split_fixed(colnames(assay_data), pattern = "_", n=3)[,1]
)
```

## Filter to Matching Phenotype Data

```{r filter_idat}
assay_data <- assay_data[,pheno_df$sample]
```

# Predict Age

## Get Beta Values

```{r age_beta}
suppressMessages({
  beta_data <- minfi::getBeta(assay_data)
})
```

## Summarize Missing Cortical Probe Values

```{r coef_summary}
brain_coef <- read.csv(file.path(reference_dir, "brain_coef.csv"), row.names = 1)
beta_coef <- beta_data[row.names(beta_data) %in% brain_coef$probe, ]
beta_missing <- is.na(beta_coef)

df_missing <- data.frame(
  probes = c(
    "CorticalClock probe count (N)",
    "CorticalClock probes in data (N)",
    "Sample count (N)",
    "Probes with Missing Values (N)",
    "Samples with Missing Values (N)",
    "Total Missing Values (N)"
  ),
  value = c(
    nrow(brain_coef),
    nrow(beta_coef),
    ncol(beta_coef),
    sum(rowSums(beta_missing) > 0),
    sum(colSums(beta_missing) > 0),
    sum(beta_missing)
  )
)

df_missing %>%
  DT::datatable(
    filter = 'top',
    style = "bootstrap",
    extensions = 'Buttons',
    options = list(
      scrollX = TRUE, 
      dom = 'Bfrtip',
      buttons = I('colvis'),
      keys = TRUE, 
      pageLength = 10
    ), 
    rownames = FALSE,
    caption = "Number of NAs in CorticalClock Probe Beta Values"
  )
```

## Run Prediction

```{r}
pheno <- CorticalClock(
  beta_data,
  pheno_df,
  paste0(reference_dir, "/"),
  "sample",
  "age"
)

pheno_df <- pheno_df %>%
  dplyr::mutate(age_prediction = as.numeric(pheno$brainpred)) %>%
  dplyr::mutate(age_truncated = ifelse(
    .data$age_prediction > 90,
    90,
    .data$age_prediction
  ))
```

# Save

```{r save, eval = TRUE}
saveRDS(assay_data, file.path(compile_dir, "organized_assay.RDS"))
saveRDS(pheno_df, file.path(compile_dir, "organized_pheno.RDS"))
write.csv(pheno_df, file.path(compile_dir, "organized_pheno.csv"))
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()
```
</details>