---
title: "Evaluate if the associations in normal, AD, and all samples are the same or not"
author:
  - David Lukacsovich^[University of Miami]
  - Lily Wang^[University of Miami]
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::html_document:
    highlight: breezedark
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up

## Load Libraries

```{r library}
suppressMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})

base_dir <- "~/TBL Dropbox/David Lukacsovich"
analysis_dir <- file.path(base_dir, "AD-Aging-brain-sample-analysis")
result_dir <- file.path(analysis_dir, "analysis-results")
code_dir <- file.path(analysis_dir, "code")
dataset_dir <- file.path(code_dir, "DATASETS")
rosmap_dir <- file.path(dataset_dir, "Rosmap")
reference_dir <- file.path(code_dir, "Reference_Files")
rna_dir <- file.path(base_dir, "coMethDMR_metaAnalysis/DNAm_RNA/data")
```

## Load Results

```{r load}
dnam_df <- read.csv(file.path(
  result_dir, "DNAm_RNA", "All_CpG_to_RNA_Association.csv")
) %>%
  dplyr::select(
    "CpG",
    "ens",
    "gene",
    "estimate",
    "pvalue",
    "fdr",
    count_all = "count_total",
    "count_norm",
    "count_ad"
  ) %>%
  dplyr::mutate(plog = -log10(.data$pvalue)) %>%
  dplyr::mutate(sig = ifelse(.data$fdr < 0.05, "signif", "neutral"))
dnam_norm_df <- read.csv(file.path(
  result_dir, "DNAm_RNA", "Normal_CpG_to_RNA_Association.csv")
) %>%
  dplyr::select(
    "CpG",
    "ens",
    "gene",
    "estimate",
    "pvalue",
    "fdr"
  ) %>%
  dplyr::mutate(plog = -log10(.data$pvalue)) %>%
  dplyr::mutate(sig = ifelse(.data$fdr < 0.05, "signif", "neutral"))
dnam_ad_df <- read.csv(file.path(
  result_dir, "DNAm_RNA", "AD_CpG_to_RNA_Association.csv")
) %>%
  dplyr::select(
    "CpG",
    "ens",
    "gene",
    "estimate",
    "pvalue",
    "fdr"
  ) %>%
  dplyr::mutate(plog = -log10(.data$pvalue)) %>%
  dplyr::mutate(sig = ifelse(.data$fdr < 0.05, "signif", "neutral"))
```

# Volcano Plots

Volcano plots can be used to plot significance vs estimate, to see if a stronger correlation corresponds to more significance. It is a good initial check to make sure that a distribution makes sense

## Define Function

```{r volcano_function}
create_volcano <- function(dnam_df, title) {
  thresh <- dnam_df %>%
    dplyr::arrange(dplyr::desc(.data$plog)) %>%
    head(n = 5) %>%
    dplyr::pull("plog") %>%
    min()
  
  plot_df <- dnam_df %>%
    dplyr::mutate(
      label = ifelse(.data$plog >= thresh, paste0(.data$CpG, "_", .data$gene), "")
    )
  
  ggplot2::ggplot(
    plot_df,
    mapping = ggplot2::aes(x = estimate, y = plog, color = sig, label = label)
  ) +
    ggplot2::geom_point(size = 1) +
    ggplot2::xlab("Correlation Estimate") +
    ggplot2::ylab("-log10(p-value)") +
    ggplot2::ggtitle(title) +
    ggrepel::geom_text_repel()
}
```


## Plot Normal Data

```{r volcano_normal}
create_volcano(dnam_norm_df, "Normal Samples")
```

## Plot AD Data

```{r volcano_ad}
create_volcano(dnam_ad_df, "AD Samples")
```

## Plot All Data

```{r volcano_all}
create_volcano(dnam_df, "All Samples")
```

Overall, the plots show the expected distributions - with a few artifacts to the side - and the top 5 most significant correlations appear to be consistent between AD and all samples.

# Correlation Plots

Next we plot the estimates and log of p-values across different methods against each other, to see how close to a linear fit that we get.

## Define Functions

```{r compare_functions}
estimates_plot <- function(df_1, df_2, xlabel, ylabel) {
  df_1 <- df_1 %>%
    dplyr::select("CpG", "ens", "gene", estimate_x = "estimate", sig_x = "sig")
  df_2 <- df_2 %>%
    dplyr::select("CpG", "ens", "gene", estimate_y = "estimate", sig_y = "sig")
  
  plot_df <- df_1 %>%
    dplyr::full_join(df_2, by = c("CpG", "ens", "gene")) %>%
    dplyr::mutate(label = paste0(.data$CpG, "_", .data$gene), to_label = FALSE) %>%
    dplyr::mutate(
      signif = paste0(.data$sig_x, "_", .data$sig_y)
    ) %>%
    dplyr::mutate(
      signif = ifelse(.data$signif == "neutral_neutral", "neither", .data$signif),
      signif = ifelse(.data$signif == "neutral_signif", ylabel, .data$signif),
      signif = ifelse(.data$signif == "signif_neutral", xlabel, .data$signif),
      signif = ifelse(.data$signif == "signif_signif", "both", .data$signif)
    ) %>%
    dplyr::mutate(
      size = ifelse(.data$signif == "neither", 0.1, 1)
    )
  
  plot_df <- rbind(
    plot_df %>%
      dplyr::filter(.data$signif == "neither"),
    plot_df %>%
      dplyr::filter(.data$signif != "neither")
  )
  
  ggplot2::ggplot(plot_df, mapping = ggplot2::aes(x = estimate_x, y = estimate_y, color = signif)) +
    ggplot2::geom_point(size = 1) +
    ggplot2::xlab(paste0(xlabel, " Samples, Correlation Estimate")) +
    ggplot2::ylab(paste0(ylabel, " Samples, Correlation Estimate")) +
    ggplot2::ggtitle("Compare Estimates")
}

plog_plot <- function(df_1, df_2, xlabel, ylabel) {
  df_1 <- df_1 %>%
    dplyr::mutate(
      plog = sign(.data$estimate) * .data$plog
    ) %>%
    dplyr::select("CpG", "ens", "gene", plog_x = "plog", sig_x = "sig")
  df_2 <- df_2 %>%
    dplyr::mutate(
      plog = sign(.data$estimate) * .data$plog
    ) %>%
    dplyr::select("CpG", "ens", "gene", plog_y = "plog", sig_y = "sig")
  
  plot_df <- df_1 %>%
    dplyr::full_join(df_2, by = c("CpG", "ens", "gene")) %>%
    dplyr::mutate(label = paste0(.data$CpG, "_", .data$gene), to_label = FALSE) %>%
    dplyr::mutate(
      signif = paste0(.data$sig_x, "_", .data$sig_y)
    ) %>%
    dplyr::mutate(
      signif = ifelse(.data$signif == "neutral_neutral", "neither", .data$signif),
      signif = ifelse(.data$signif == "neutral_signif", ylabel, .data$signif),
      signif = ifelse(.data$signif == "signif_neutral", xlabel, .data$signif),
      signif = ifelse(.data$signif == "signif_signif", "both", .data$signif)
    ) %>%
    dplyr::mutate(
      size = ifelse(.data$signif == "neither", 0.1, 1)
    )
  
  plot_df <- rbind(
    plot_df %>%
      dplyr::filter(.data$signif == "neither"),
    plot_df %>%
      dplyr::filter(.data$signif != "neither")
  )
  
  ggplot2::ggplot(plot_df, mapping = ggplot2::aes(x = plog_x, y = plog_y, color = signif)) +
    ggplot2::geom_point(size = 1) +
    ggplot2::xlab(paste0(xlabel, " Samples, Correlation -log10(p-value)")) +
    ggplot2::ylab(paste0(ylabel, " Samples, Correlation -log10(p-value)")) +
    ggplot2::ggtitle("Compare Significance")
}
```

## Compare Estimates

### Normal vs AD

```{r estimate_norm_ad}
estimates_plot(dnam_norm_df, dnam_ad_df, "Normal", "AD")
```

### Normal vs All

```{r estimate_norm_all}
estimates_plot(dnam_norm_df, dnam_df, "Normal", "All")
```

### AD vs All

```{r estimate_ad_all}
estimates_plot(dnam_ad_df, dnam_df, "AD", "All")
```

## Compare Significance

Compare log10 of p-values, adjusted by signs of estimates

### Normal vs AD

```{r signif_norm_ad}
plog_plot(dnam_norm_df, dnam_ad_df, "Normal", "AD")
```

### Normal vs All

```{r signif_norm_all}
plog_plot(dnam_norm_df, dnam_df, "Normal", "All")
```

### AD vs All

```{r signif_ad_all}
plog_plot(dnam_ad_df, dnam_df, "AD", "All")
```

# Calculate Correlation Scores

While plots help with visualization, it is useful to have concrete correlation scores to look at as well.

## Define Functions

```{r corr_functions}
compare_estimates <- function(df_1, df_2, filter_sig = FALSE) {
  est_df <- df_1 %>%
    dplyr::select("CpG", "ens", estimate_x = "estimate", sig_x = "sig") %>%
    dplyr::full_join(
      df_2 %>%
        dplyr::select("CpG", "ens", estimate_y = "estimate", sig_y = "sig")
    )
  
  if (filter_sig) {
    est_df <- est_df %>%
      dplyr::filter(.data$sig_x == "signif" | .data$sig_y == "signif")
  }
  
  corr.1 <- cor.test(est_df$estimate_x, est_df$estimate_y, method = "spearman")
  corr.2 <- cor.test(est_df$estimate_x, est_df$estimate_y, method = "pearson")
  
  list(corr.spear = corr.1, corr.pear = corr.2, count = nrow(est_df))
}

get_estimate_scores <- function(dnam_df, dnam_norm_df, dnam_ad_df, filter_sig = FALSE) {
  norm_ad <- compare_estimates(dnam_norm_df, dnam_ad_df, filter_sig = filter_sig)
  corr.1 <- norm_ad$corr.spear
  corr.2 <- norm_ad$corr.pear
  
  norm_all <- compare_estimates(dnam_norm_df, dnam_df, filter_sig = filter_sig)
  corr.3 <- norm_all$corr.spear
  corr.4 <- norm_all$corr.pear
  
  all_ad <- compare_estimates(dnam_df, dnam_ad_df, filter_sig = filter_sig)
  corr.5 <- all_ad$corr.spear
  corr.6 <- all_ad$corr.pear
  
  data.frame(
    comparison = c(
      "Normal vs AD",
      "Normal vs AD",
      "Normal vs All",
      "Normal vs All",
      "AD vs All",
      "AD vs All"
    ),
    test = c(
      "Spearman",
      "Pearson",
      "Spearman",
      "Pearson",
      "Spearman",
      "Pearson"
    ),
    estimate = c(
      corr.1$estimate,
      corr.2$estimate,
      corr.3$estimate,
      corr.4$estimate,
      corr.5$estimate,
      corr.6$estimate
    ),
    pvalue = c(
      corr.1$p.value,
      corr.2$p.value,
      corr.3$p.value,
      corr.4$p.value,
      corr.5$p.value,
      corr.6$p.value
    ),
    n = c(
      norm_ad$count,
      norm_ad$count,
      norm_all$count,
      norm_all$count,
      all_ad$count,
      all_ad$count
    )
  )
}

get_plog_scores <- function(dnam_df, dnam_norm_df, dnam_ad_df, filter_sig = FALSE) {
  get_estimate_scores(
    dnam_df %>% dplyr::mutate(estimate = sign(.data$estimate) * .data$plog),
    dnam_norm_df %>% dplyr::mutate(estimate = sign(.data$estimate) * .data$plog),
    dnam_ad_df %>% dplyr::mutate(estimate = sign(.data$estimate) * .data$plog),
    filter_sig = filter_sig
  )
}
```

## Get Correlation Scores

```{r corr_eval}
est_df <- get_estimate_scores(dnam_df, dnam_norm_df, dnam_ad_df, filter_sig = FALSE)
plog_df <- get_plog_scores(dnam_df, dnam_norm_df, dnam_ad_df, filter_sig = FALSE)

est_sig_df <- get_estimate_scores(dnam_df, dnam_norm_df, dnam_ad_df, filter_sig = TRUE)
plog_sig_df <- get_plog_scores(dnam_df, dnam_norm_df, dnam_ad_df, filter_sig = TRUE)
```

## Display Correlation of DNAm to RNA associations

### Estimate Correlations

```{r corr_disp_est}
est_df
```

### Significance Correlations

```{r corr_disp_plog}
plog_df
```

## Display Correlation of Significant DNAm to RNA associations

### Estimate Correlations

```{r corr_disp_sig_est}
est_sig_df
```

### Significance Correlations

```{r corr_disp_sig_plog}
plog_sig_df
```

# Tabulate Directions

We want to see how many comparisons - both significant and not - shared directions in different sample groups

## Define Functions

```{r tabular_functions}
get_tables <- function(df_1, df_2) {
  comp_df <- df_1 %>%
    dplyr::mutate(direction = ifelse(.data$estimate > 0, "+", "-")) %>%
    dplyr::select("CpG", "ens", sign_x = "direction", sig_x = "sig") %>%
    dplyr::full_join(
      df_2 %>%
        dplyr::mutate(direction = ifelse(.data$estimate > 0, "+", "-")) %>%
        dplyr::select("CpG", "ens", sign_y = "direction", sig_y = "sig")
    ) %>%
    dplyr::mutate(sig = ((.data$sig_x == "signif") | (.data$sig_y == "signif")))
  
  list(
    all = comp_df %>%
      dplyr::select("sign_x", "sign_y") %>%
      table(),
    base = comp_df %>%
      dplyr::filter(!.data$sig) %>%
      dplyr::select("sign_x", "sign_y") %>%
      table(),
    signif = comp_df %>%
      dplyr::filter(.data$sig) %>%
      dplyr::select("sign_x", "sign_y") %>%
      table()
  )
}
```


## Normal vs AD

```{r tabular_norm_ad}
res <- get_tables(dnam_norm_df, dnam_ad_df)

print("All Comparisons (x = Normal, y = AD)")
res$all
print("Non-Signif Comparisons (x = Normal, y = AD)")
res$base
print("Signif Comparisons (x = Normal, y = AD)")
res$signif
```

## Normal vs All

```{r tabular_norm_all}
res <- get_tables(dnam_norm_df, dnam_df)

print("All Comparisons (x = Normal, y = All Samples)")
res$all
print("Non-Signif Comparisons (x = Normal, y = All Samples)")
res$base
print("Signif Comparisons (x = Normal, y = All Samples)")
res$signif
```

## AD vs All

```{r tabular_ad_all}
res <- get_tables(dnam_ad_df, dnam_df)

print("All Comparisons (x = AD, y = All Samples)")
res$all
print("Non-Signif Comparisons (x = AD, y = All Samples)")
res$base
print("Signif Comparisons (x = AD, y = All Samples)")
res$signif
```

# Save Signifs

Make a file containing only CpG-gene comparisons that are significant in at least 1 method

```{r}
group_df <- dnam_df %>%
  dplyr::select(
    "CpG",
    "ens",
    "gene",
    estimate_all = "estimate",
    pvalue_all = "pvalue",
    fdr_all = "fdr",
    "count_all",
    "count_norm",
    "count_ad"
  ) %>%
  dplyr::full_join(
    dnam_norm_df %>%
      dplyr::select(
        "CpG",
        "ens",
        "gene",
        estimate_norm = "estimate",
        pvalue_norm = "pvalue",
        fdr_norm = "fdr"
      ),
    by = c("CpG", "ens", "gene")
  ) %>%
  dplyr::full_join(
    dnam_ad_df %>%
      dplyr::select(
        "CpG",
        "ens",
        "gene",
        estimate_ad = "estimate",
        pvalue_ad = "pvalue",
        fdr_ad = "fdr"
      ),
    by = c("CpG", "ens", "gene")
  ) %>%
  dplyr::select(
    "CpG",
    "ens",
    "gene",
    "estimate_all",
    "estimate_norm",
    "estimate_ad",
    "pvalue_all",
    "pvalue_norm",
    "pvalue_ad",
    "fdr_all",
    "fdr_norm",
    "fdr_ad",
    "count_all",
    "count_norm",
    "count_ad"
  ) %>%
  dplyr::filter(
    (.data$fdr_all <= 0.05) |
      (.data$fdr_norm <= 0.05) |
      (.data$fdr_ad <= 0.05)
  )

write.csv(group_df, file = file.path(
  result_dir, "DNAm_RNA", "CpG_to_RNA_Significant_Comparisons.csv"
))
```

# Evaluate DMR Signifs

We want to look at and evaluate DMR signifs as well

## Load DMR Results

```{r load_dmr}
dmr_df <- read.csv(file.path(
  result_dir, "DNAm_RNA", "All_DMR_to_RNA_Association.csv"
)) %>%
  dplyr::select(
    "DMR",
    "combp_DMR",
    "ens",
    "gene",
    "estimate",
    "pvalue",
    "fdr",
    count_all = "count_total",
    "count_norm",
    "count_ad"
  ) %>%
  dplyr::mutate(plog = -log10(.data$pvalue)) %>%
  dplyr::mutate(sig = ifelse(.data$fdr < 0.05, "signif", "neutral")) %>%
  dplyr::mutate(CpG = .data$combp_DMR)

dmr_norm_df <- read.csv(
  file.path(
    result_dir, "dnam_RNA", "Normal_DMR_to_RNA_Association.csv"
  )
) %>%
  dplyr::select(
    "DMR",
    "combp_DMR",
    "ens",
    "gene",
    "estimate",
    "pvalue",
    "fdr"
  ) %>%
  dplyr::mutate(plog = -log10(.data$pvalue)) %>%
  dplyr::mutate(sig = ifelse(.data$fdr < 0.05, "signif", "neutral")) %>%
  dplyr::mutate(CpG = .data$combp_DMR)

dmr_ad_df <- read.csv(
  file.path(
    result_dir, "dnam_RNA", "AD_DMR_to_RNA_Association.csv"
  )
) %>%
  dplyr::select(
    "DMR",
    "combp_DMR",
    "ens",
    "gene",
    "estimate",
    "pvalue",
    "fdr"
  ) %>%
  dplyr::mutate(plog = -log10(.data$pvalue)) %>%
  dplyr::mutate(sig = ifelse(.data$fdr < 0.05, "signif", "neutral")) %>%
  dplyr::mutate(CpG = .data$combp_DMR)
```

## Get Correlation Scores

```{r corr_eval_dmr}
est_df <- get_estimate_scores(dmr_df, dmr_norm_df, dmr_ad_df, filter_sig = FALSE)
plog_df <- get_plog_scores(dmr_df, dmr_norm_df, dmr_ad_df, filter_sig = FALSE)

est_sig_df <- get_estimate_scores(dmr_df, dmr_norm_df, dmr_ad_df, filter_sig = TRUE)
plog_sig_df <- get_plog_scores(dmr_df, dmr_norm_df, dmr_ad_df, filter_sig = TRUE)
```

### Display Correlation of DMR to RNA associations

#### Estimate Correlations

```{r corr_disp_est_dmr}
est_df
```

#### Significance Correlations

```{r corr_disp_plog_dmr}
plog_df
```

### Display Correlation of Significant DMR to RNA associations

#### Estimate Correlations

```{r corr_disp_sig_est_dmr}
est_sig_df
```

#### Significance Correlations

```{r corr_disp_sig_plog_dmr}
plog_sig_df
```

## Merge and Filter DMR Filters

```{r merge_dmr}
group_df <- dmr_df %>%
  dplyr::select(
    "DMR",
    "combp_DMR",
    "ens",
    "gene",
    estimate_all = "estimate",
    pvalue_all = "pvalue",
    fdr_all = "fdr",
    "count_all",
    "count_norm",
    "count_ad"
  ) %>%
  dplyr::full_join(
    dmr_norm_df %>%
      dplyr::select(
        "DMR",
        "combp_DMR",
        "ens",
        "gene",
        estimate_norm = "estimate",
        pvalue_norm = "pvalue",
        fdr_norm = "fdr"
      ),
    by = c("DMR", "combp_DMR", "ens", "gene")
  ) %>%
  dplyr::full_join(
    dmr_ad_df %>%
      dplyr::select(
        "DMR",
        "combp_DMR",
        "ens",
        "gene",
        estimate_ad = "estimate",
        pvalue_ad = "pvalue",
        fdr_ad = "fdr"
      ),
    by = c("DMR", "combp_DMR", "ens", "gene")
  ) %>%
  dplyr::select(
    "DMR",
    "combp_DMR",
    "ens",
    "gene",
    "estimate_all",
    "estimate_norm",
    "estimate_ad",
    "pvalue_all",
    "pvalue_norm",
    "pvalue_ad",
    "fdr_all",
    "fdr_norm",
    "fdr_ad",
    "count_all",
    "count_norm",
    "count_ad"
  ) %>%
  dplyr::filter(
    (.data$fdr_all <= 0.05) |
      (.data$fdr_norm <= 0.05) |
      (.data$fdr_ad <= 0.05)
  )
```

## Save DMR Results

```{r save_dmr}
write.csv(group_df, file = file.path(
  result_dir, "DNAm_RNA", "DMR_to_RNA_Significant_Comparisons.csv")
)
```

# Evaluate Merged Correlation

## Merge Data Points

```{r merge_dmr_dnam}
all_df <- rbind(
  dnam_df %>% dplyr::select("CpG", "ens", "estimate", "plog", "sig"),
  dmr_df %>% dplyr::select("CpG", "ens", "estimate", "plog", "sig")
)

norm_df <- rbind(
  dnam_norm_df %>% dplyr::select("CpG", "ens", "estimate", "plog", "sig"),
  dmr_norm_df %>% dplyr::select("CpG", "ens", "estimate", "plog", "sig")
)

ad_df <- rbind(
  dnam_ad_df %>% dplyr::select("CpG", "ens", "estimate", "plog", "sig"),
  dmr_ad_df %>% dplyr::select("CpG", "ens", "estimate", "plog", "sig")
)
```

## Calculate Scores

```{r merge_scores}
est_df <- get_estimate_scores(all_df, norm_df, ad_df, filter_sig = FALSE)
plog_df <- get_plog_scores(all_df, norm_df, ad_df, filter_sig = FALSE)

est_sig_df <- get_estimate_scores(all_df, norm_df, ad_df, filter_sig = TRUE)
plog_sig_df <- get_plog_scores(all_df, norm_df, ad_df, filter_sig = TRUE)
```

## Show Correlation Tabels

### Display Correlation of DMR to RNA associations

#### Estimate Correlations

```{r corr_disp_est_merge}
est_df
```

#### Significance Correlations

```{r corr_disp_plog_merge}
plog_df
```

### Display Correlation of Significant DMR to RNA associations

#### Estimate Correlations

```{r corr_disp_sig_est_merge}
est_sig_df
```

#### Significance Correlations

```{r corr_disp_sig_plog_merge}
plog_sig_df
```

# Session Information

<details>
  <summary>**Session Info**</summary>
```{r session}
sessionInfo()
```

